Measuring causal influence with\\ back-to-back regression

### 2019-05-23 paper/introduction.tex
Updates

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart_neuroimage_2015, king2018encoding}.

>>>>

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}.

### 2019-05-23 paper/meg_exps.tex
Updates

The raw MEG data was bandpass-filtered between 0.1 and 40Hz using MNE-Python
default parameters \citep{gramfort2013meg, gramfort2014mne}. Specifically, we used a zero-phase finite impulse
response filter (FIR) with a Hamming window and with transition bands of 0.1Hz
and 10Hz for the low and high cut-off frequencies respectively.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/ridgecv_baseline_result.pdf}
  \caption{Ridge regression baseline. All the coefficients appear to be
  linearly correlated, unlike our method in Figure \ref{fig:megresult}.}
  \label{fig:ridgebaselineresult}
\end{figure}

>>>>

The raw MEG data was bandpass-filtered between 0.1 and 40Hz using MNE-Python
default parameters \citep{gramfort2013meg, gramfort2014mne}. Specifically, we used a zero-phase finite impulse
response filter (FIR) with a Hamming window and with transition bands of 0.1Hz
and 10Hz for the low and high cut-off frequencies.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/ridgecv_baseline_result.pdf}
  \caption{Regression baseline. Coefficients appear to be
  linearly correlated, unlike our method Fig. \ref{fig:megresult}.}
  \label{fig:ridgebaselineresult}
\end{figure}

### 2019-05-23 paper/paper.tex
Updates

% TODO: Talk about cause-effect methods, trace method, half-sibling
% TODO: Cite Roweis component analysis methods paper
% TODO: read quantifying causal influences
% TODO: review abstract
% TODO: review conclusion

@@

We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
%
In summary, our model can be formalized as:
%
\begin{equation}
    Y = (XE + N)F,\label{eq:model}
\end{equation}
%
where $N$ is a centered, unobserved noise variable.
%
While the triplet of variables $X$ and $N$ are independent, we allow each of them to have a general covariance matrix.
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{efron1992bootstrap}.
%
After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.

We summarize the B2BR procedure in Algorithm~\ref{algorithm:b2br}.
%
The rest of this section provides a theoretical guarantee on the correctness of B2BR, and an optional post-processing step to binarize the estimated causal influence matrix.

@@

%%% Next, let $k$ be the number of ones in the binary diagonal causal influence matrix $E$.
%%% %
%%% That is, $\text{tr}(E) = k$.
%%% %
%%% Without loss of generality, assume that these are the first $k$ variables in $X$.
%%% %
%%% Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
%%% %
%%% Let $K \in \mathbb{R}^{k \times d_{x}}$ be the diagonal matrix obtained by deleting the bottom $d_{x} - k$ rows of $E$.
%%% %
%%% %
%%% First, we factorize $(XEF)^\dagger = ((XK^\top)(KF))^\dagger = (KF)^\dagger (XK^\top)^\dagger$.
%%% %
%%% Second, since $XK^\top$ is full rank, we write $(XK^\top)^\dagger = ((XK^\top)^\top (XK^\top))^{-1}(XK^\top)^\top$.
%%% %
%%% Then, $(XEF)^\dagger X = (KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X$.
%%% 
%%% %Recall the rank factorisation property of the pseudo-inverse \citep{ben2003generalized}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
%%% %
%%% % Using this property on the minimal norm solution $\hat G$ yields
%%% % \begin{equation}
%%% % \begin{aligned}
%%% % (XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
%%% % &=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
%%% % &=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
%%% % \end{aligned}
%%% % \end{equation}
%%% 
%%% Write the covariance of $X$ as $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, with the block $\Sigma_1$ being $k\times k$. Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{c|c}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right) =: S$.
%%% 
%%% 
%%% Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
%%% 
%%% Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$. Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).
%%% 
%%% For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$. (Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$). The square norm is therefore $\lambda^2 Var(N_{active})$.
%%% 
%%% Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$. Its least value is attained for $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N_{active})}$. Assuming X and N have the same variance over all features (this can always be guaranteed), active subscript can be dropped, and this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
%%% 
%%% The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} (KF)^\dagger S$.
%%% 
%%% 
%%% Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
%%% 
%%% ADD AGAIN:
%%% 
%%% Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$. If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.

@@

\section{Related work}

@@

>>>>

@@

We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
%
In summary, our model can be formalized as:
%
\begin{equation}
    Y = (XE + N)F,\label{eq:model}
\end{equation}
%
where $N$ is a centered, unobserved noise variable.
%
While the triplet of variables $X$ and $N$ are independent, we allow each of them to have a general covariance matrix.
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2B) is to estimate the matrix of causal influences $E$.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2B is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{efron1992bootstrap}.
%
After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$.

We summarize the B2B procedure in Algorithm~\ref{algorithm:b2br}.
%
The rest of this section provides a theoretical guarantee on the correctness of B2B, and an optional post-processing step to binarize the estimated causal influence matrix.

@@

%%%%%% Next, let $k$ be the number of ones in the binary diagonal causal influence matrix $E$.
%%%%%% %
%%%%%% That is, $\text{tr}(E) = k$.
%%%%%% %
%%%%%% Without loss of generality, assume that these are the first $k$ variables in $X$.
%%%%%% %
%%%%%% Assume that $XEF$ is of rank $k$, which means that $F$ and $X$ have full rank on the space spanned by the active elements in $E$.
%%%%%% %
%%%%%% Let $K \in \mathbb{R}^{k \times d_{x}}$ be the diagonal matrix obtained by deleting the bottom $d_{x} - k$ rows of $E$.
%%%%%% %
%%%%%% %
%%%%%% First, we factorize $(XEF)^\dagger = ((XK^\top)(KF))^\dagger = (KF)^\dagger (XK^\top)^\dagger$.
%%%%%% %
%%%%%% Second, since $XK^\top$ is full rank, we write $(XK^\top)^\dagger = ((XK^\top)^\top (XK^\top))^{-1}(XK^\top)^\top$.
%%%%%% %
%%%%%% Then, $(XEF)^\dagger X = (KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X$.
%%%%%% 
%%%%%% %Recall the rank factorisation property of the pseudo-inverse \citep{ben2003generalized}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
%%%%%% %
%%%%%% % Using this property on the minimal norm solution $\hat G$ yields
%%%%%% % \begin{equation}
%%%%%% % \begin{aligned}
%%%%%% % (XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
%%%%%% % &=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
%%%%%% % &=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
%%%%%% % \end{aligned}
%%%%%% % \end{equation}
%%%%%% 
%%%%%% Write the covariance of $X$ as $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, with the block $\Sigma_1$ being $k\times k$. Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{c|c}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right) =: S$.
%%%%%% 
%%%%%% 
%%%%%% Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
%%%%%% 
%%%%%% Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$. Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).
%%%%%% 
%%%%%% For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$. (Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$). The square norm is therefore $\lambda^2 Var(N_{active})$.
%%%%%% 
%%%%%% Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$. Its least value is attained for $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N_{active})}$. Assuming X and N have the same variance over all features (this can always be guaranteed), active subscript can be dropped, and this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
%%%%%% 
%%%%%% The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} (KF)^\dagger S$.
%%%%%% 
%%%%%% 
%%%%%% Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
%%%%%% 
%%%%%% ADD AGAIN:
%%%%%% 
%%%%%% Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$. If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.
%%%=======
%%%
%%%Let $X^\top X = \left(\begin{array}{c|c}\Sigma_{1} & \Sigma_{2} \\\hline \Sigma_{2} & \Sigma_{3}\end{array}\right)$, the top left block being $k\times k$. Then, $(KX^\top X K^\top)=\Sigma_{1}$ and $(KX^\top XK^\top)^{-1} KX^\top X=\left(\begin{array}{cc}I_{k} &  \Sigma_{1}^{-1} \Sigma_{2}\end{array}\right)$, call it S.
%%%
%%%Thus, in the absence of noise, and if F and X have full rank on the subspace spanned by E, we have $G\hat=(XEF)^\dagger X=(KF)^\dagger S$. If the covariance of X is block diagonal, that is if inactive features and active features are not correlated (if the causal features do not serve as confounders for non causal ones), then S=K, and we are retrieving $(KF)^\dagger K$.
%%%
%%%Under centred noise, and if the covariance of NF has low condition number (a weak assumption given that N is noise), we have $\left \| NFG\right \| ^2 \approx Var(NF) \left \| G\right \| ^2$, and minimisation of $\left \| X-XEFG\right\| ^2  + \left \| NFG\right \| ^2$ will recover a scaled version of the noiseless solution, $\lambda (KF)^\dagger S$, $\lambda$ a positive number below one.
%%%
%%%Replacing in the left part of (ref), we have $X-XEFG=X- XK^\top KF \lambda (KF)^\dagger S= X - \lambda XE=X(I-\lambda E)$. Its square norm is $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})$ ($Var(X_{.})$ denoting the variance of X over active and inactive features).
%%%
%%%For the right part, we have $NFG=NF \lambda (KF)^\dagger S = NK^\top (KF+(I-K)F) \lambda  (KF)^\dagger S = \lambda NES=  \lambda NE$. (Since F is full rank over E, we have $(KF)^\dagger (I-K)F= 0$). The square norm is therefore $\lambda^2 Var(N_{active})$.
%%%
%%%Thus, with centred and random noise, and with F full rank over E, we are minimising $(1-\lambda)^2 Var(X_{active}) + Var(X_{inactive})+\lambda^2 Var(N_{active})$. Its least value is attained for $\lambda = \frac{Var(X_{active})}{Var(X_{active})+Var(N_{active})}$. Assuming X and N have the same variance over all features (this can always be guaranteed), active subscript can be dropped, and this simplifies to $\lambda= \frac {Var(X)}{Var(X)+ Var(N)}=\frac{1}{1+nsr}$ nsr the noise to signal ratio, Var(N) divided by Var(X).
%%%
%%%The first step of the back to back regression therefore retrieves $\frac{1}{1+nsr} S (KF)^\dagger$.
%%%
%%%The second step is straightforward, replacing Y, we have
%%%\begin{equation}
%%%\begin{aligned}
%%%\arg \min_H \left \| \hat YG - XH \right \|^2 &=\arg  \min_H \left \| (XE + N)F \hat G - XH \right \|^2 \\
%%%&=\arg \min_H \left \| X(EF \hat G - H) \right \| ^2 + \left \| NF\hat G \right \| ^2\\
%%%&= \arg \min_H \left \| X(EF \hat G - H) \right \| ^2\\
%%%&= EF \hat G
%%%\end{aligned}
%%%\end{equation}
%%%
%%%Thus, in the linear case, back to back regression retrieves $\hat H = \frac{1}{1+nsr} (K^\top K F)(KF)^{\dagger} S = \frac{1}{1+nsr} E$.
%%%>>>>>>> 4d110522fbb65b4da60327da2627241785a06113

@@

\section{Related work}
% scholkopf2016modeling

@@

B2BR is also related, but different, from usual causal discovery algorithms \citep{peters2017elements} that hypothesize on the direction of causality, or from measures of causal influence requiring time information \citep{granger1969investigating, janzing2013quantifying}.

### 2019-05-23 paper/synthetic.tex
Updates

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{synthetic.pdf}
    \caption{Results of synthetic experiments. Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration, averaged over $20$ random seeds.}
    \label{table:synthetic}
\end{figure}

>>>>

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{synthetic.pdf}
    \caption{Results of synthetic experiments. Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration, $20$ random seeds.}
    \label{table:synthetic}
\end{figure}
